A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-05 doc/popl2013/langdef.tex

\newcommand{\oton}[1]{{#1}_1,\ldots,{#1}_n}
\newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}
\newcommand{\dplus}{\text{+\!+}}
\newcommand{\llbracket}{[\![}
\newcommand{\rrbracket}{]\!]}

@@

$T$ & ::= &  & types \\
  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types \\
  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple \\
  &$\mid$& $T@(\oton{r})$ & pointer \\
  &$\mid$& $\text{coloring}(r)$ & region coloring \\
  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship \\
  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions \\
\\
$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints \\
$\omega$ & ::= & $r_1 \leq r_2$ & subregion \\
  &$\mid$& $r_1 * r_2$ & disjointness \\
\\
$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges \\
$\phi$ & ::= & reads$(r)$ $\;\;\;\mid\;\;\;$ writes$(r)$ $\;\;\;\mid\;\;\;$ reduces$(id,r)$ & \\
\\
$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes \\
$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & \\
\\
$v$ & ::= & & values \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values \\
  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple \\
  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address \\
  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring \\
  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & region relationship instance \\
\\
$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
\\
$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
\\
$e$ & ::= & & expressions \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
  &$\mid$& $id$ &  \\
  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
  &$\mid$& $e_1 + e_2$ & integer operations \\
  &$\mid$& $e_1 < e_2$ & comparison operations \\
  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\

@@

\section{Legion Core Language}

\subsection{Language Definition}

@@

\forall i, reduces(id, r_i) \in \Phi^*

@@

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces(id, r_j) \in \Phi^* \Rightarrow reduces(id, r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), \Phi', Q' \rightarrow T_r \Rightarrow reduces(id, r) \in \Phi^*
\end{array}$
}
\caption{Privilege and Constraint Closure}
\label{fig:closure}
\end{figure}

@@

\begin{center}
\begin{tabular}{l@{ }l@{ }l}
$M \vdash E : \Phi$ & $\Leftrightarrow$ & $\forall \epsilon\text { in } E.$ \\
& & $(\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge$ \\
& & $(\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
& & $(\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces(id, r) \in \Phi)$
\end{tabular}
\end{center}

@@

\begin{lem}
\label{lemma:phiclosure}
$M \vdash E : \Phi^*$ if ant only if $M \vdash E : \Phi$.
\end{lem}

@@

Figure~\ref{fig:validinterleave} gives the precise definition of the {\em valid\_interleave}
predicate, but there are a lot of moving parts, so description in ``plain'' English may help.
Consider a set of constituent traces $\oton{E}$.  Let the {\em exclusive location set} $L_{excl}$ be
the set of all locations that appear in at least one operation with coherence mode $excl$ in at least
one constituent trace, excepting any locations in the clobber set $C$.  Let the {\em atomic location
set} $L_{atomic}$ similarly be the set of locations that appear in at least one operation with 
coherence mode $atomic$, but no operations with coherence mode $excl$, again excepting any location
in the clobber set.  A memory trace $E'$ is considered a valid interleaving of $\oton{E}$ if all of
the following are true:
\begin{enumerate}
\item $E'$ is an interleaving of $\oton{E}$.
\item For every read in $E'$ with a location in $L_{excl}$, the value read is what would be in the
 store at that location if all the previous operations of $E'$ were applied to the original
store $S$.
\item For every read in $E_i$ with a location in $L_{excl}$, the value read is what would be in the
store at that location if all preceding constituent
traces (i.e. $E_1, \ldots, E_{i-1}$) were applied to the store $S$, followed by the operations from
$E_i$ that precede the read.
\item For every location $l$ in $L_{excl}$, the value
at that location after the trace $E'$ is applied to $S$ must match the value that would be in
the store if each of $E_i$ were successively applied to $S$.
\item For every read in $E'$ with a location in $L_{atomic}$ for which a write operation to the same
location appears earlier in $E'$, the value read is what would be in the store if all the previous
operations of $E'$ were applied to the initial store $S$.
\item There is some permutation $\oton{\pi}$ of the integers $1,\ldots,n$ such that, for every read
in $E_{\pi_i}$ with a location in $L_{atomic}$, the value read is what would be in the store at that
location if the traces $E_{\pi_1}, \ldots E_{\pi_{i-1}}$ were applied to $S$, following by the operations 
from $E_{\pi_i}$ that precede the read.
\item Using the same permutation $\oton{\pi}$, for every location in $L_{atomic}$, the value at that
location after the trace $E'$ is applied to $S$ must match the value that would be in the store
if each of $E_{\pi_i}$ were successively applied to $S$.
\end{enumerate}

@@

To summarize this, we define a {\em noninterference} operator on two effects that have been
annotated with which constituent trace they came from as follows:
$$
\begin{array}{l@{ }l}
op_1^{E_1}(l_1, c_1,& v_1)\ \#\ op_2^{E_2}(l_2, c_2, v_2) \Leftrightarrow E_1 \not= E_2\ \wedge \\
& \big( l_1 \not= l_2 \vee (op_1 = read \wedge op_2 = read)\ \vee \\
& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2) \big)
\end{array}
$$

\begin{lem}
\label{lem:nonintswap}
Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \# \epsilon_2$.
Then,
$$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$$
if and only if
$$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$$
\end{lem}

\begin{proof}
As before, we consider the three conditions needed to be a valid interleaving.  We describe the
implication in the ``forward'' direction, but the arguments are all symmetric.
\begin{enumerate}
\item If $E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b$ is an interleaving of $\oton{E}$,
the proof must first ``pop off'' all of $E'_a$, leaving $\epsilon_1 \dplus \epsilon_2 \dplus E'_b$ for
the interleaved subtrace.  The next two steps must pop $\epsilon_1$ and $\epsilon_2$, leaving $E'_b$.
They cannot be in the same subtrace because $\epsilon_1 \# \epsilon_2$, so we could also pop
$\epsilon_2$ and then $\epsilon_1$, resulting in a proof that $E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b$ is also an interleaving of the constituent traces.
\item For coherence, we must show the value read (if any) by $\epsilon_1$ does not change, the value
read (if any) by $\epsilon_2$ does not change, and that any read in $E'_b$ sees no change.  With the
help of Lemma~\ref{lemma:applychain}, we can eliminate the common $E'_a$ from all sequences by
using $S' = apply(S, E'_a)$.  Similarly to show that $apply(S', \epsilon_1 \dplus \epsilon_2 \dplus E) = apply(S', \epsilon_2 \dplus \epsilon_1 \dplus E)$ for any $E$, we need only to show that it holds
for $E = []$.  The proof is now reduced to showing the following:
\begin{enumerate}
\item $\epsilon_1 = read(l, c, v) \Rightarrow S'(l) = apply(S',\epsilon_2)(l)$ - since $\epsilon_1$
is a read, either $\epsilon_2$ must be a read (in which case $apply(S, \epsilon_2) = S'$) or must be
to a different location, so the result of applying just it to $S'$ cannot change the value in 
location $l$.
\item $\epsilon_2 = read(l, c, v) \Rightarrow apply(S', \epsilon_1)(l) = S'(l)$ - this is the same
as above, with $\epsilon_1$ and $\epsilon_2$ switched.
\item $apply(S', \epsilon_1 \dplus \epsilon_2) = apply(S', \epsilon_2 \dplus \epsilon_1)$ - we must
show that the two store operations $apply(\bullet, \epsilon_1)$ and $apply(\bullet, \epsilon_2)$
commute.  Modifications to two different locations commute, as do two reads (which make no change
to the store).  Finally, two reductions using the same function to the same location also commute.
\end{enumerate}
\item The proof that sequential equivalence is maintained comes down to showing the final store
is identical after the swap of the two operations, and this is merely a special case of the third
piece of the coherence proof.
\end{enumerate}
\end{proof}

We now extend the noninterference operator to entire memory traces, providing a way to say that
no operation in a trace interferes with any operation in other:
$$E_1 \# E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \# \epsilon_2$$
and observe that this property permits arbitrary ordering (i.e. parallel execution) of the 
corresponding subexpressions.

\begin{lem}
\label{lem:noninteffects}
Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \# E_j$
for every $1 \leq i < j \leq n$.  Then any trace $E'$ satisfying:
$$any\_interleave(E', \oton{E})$$
is a valid interleaving.
\end{lem}

@@

The swapping algorithm as follows: Associate with each operation in $E'$ the index of that operation
in $E_1 \dplus \ldots \dplus E_n$.  Any consectutive pair of operations $\epsilon_1$ and $\epsilon_2$
for which $index(\epsilon_1) > index(\epsilon_2)$ is a {\em misordered pair}.  For any such pair,
$\epsilon_1$ and $\epsilon_2$ will be from different constituent traces (if they were from the same
trace, their misordering would mean that $E'$ was not an interleaving of $\oton{E}$).  Every pair of
traces is noninterfering, so $\epsilon_1 \# \epsilon_2$ and they can be swapped while preserving
validity of the trace.  At most $n^2/2$ swaps are needed to eliminate all misordered pairs.
\end{proof}

Our next step is to extend noninterference to the privileges used in the type system.  We will make
use of the region mapping $M$ to translate the logical regions named in privileges into sets of
locations (i.e. physical regions), and define noninterference between individual privileges as
follows:
\begin{center}
\begin{tabular}{llc}
\multicolumn{1}{c}{$\phi_1$} & \multicolumn{1}{c}{$\phi_2$} & $\phi_1 \#_M \phi_2$? \\
\multirow{3}{*}{$reads(r_1)$} & $reads(r_2)$ & true \\
& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $reduces(id_2,r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
\cline{1-1}
\multirow{3}{*}{$writes(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $reduces(id_2,r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
\cline{1-1}
\multirow{3}{*}{$reduces(id_1r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
& $reduces(id_2,r_2)$ & $\begin{array}{c}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$ \\
\end{tabular}
\end{center}
and noninterference between sets of privileges in the same way as memory traces:
$$\Phi_1 \#_M \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \#_M \phi_2$$

@@

\begin{thm}
Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
Let $M$ be a region mapping, $L$ be local value mapping,
$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
$$
\begin{array}{l}
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$$
If $\Phi_i \#_M \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
$\oton{e}$ will result in a valid interleaving of memory operations.
\end{thm}

@@

TODO: talk about atomic case

>>>>

\newcommand{\oton}[1]{{#1}_1,\ldots,{#1}_n}
\newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}
\newcommand{\dplus}{\text{+\!+}}
\newcommand{\llbracket}{[\![}
\newcommand{\rrbracket}{]\!]}
\makeatletter
\newcommand{\nonint}[1][]{\#\@ifmtarg{#1}{}{_{\!\!#1}}}
\makeatother

@@

$T$ & ::= &  & types \\
  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types \\
  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple \\
  &$\mid$& $T@(\oton{r})$ & pointer \\
  &$\mid$& $\text{coloring}(r)$ & region coloring \\
  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship \\
  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions \\
\\
$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints \\
$\omega$ & ::= & $r_1 \leq r_2$ & subregion \\
  &$\mid$& $r_1 * r_2$ & disjointness \\
\\
$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges \\
$\phi$ & ::= & reads$(r)$ $\;\;\;\mid\;\;\;$ writes$(r)$ $\;\;\;\mid\;\;\;$ $\text{reduces}_{id}(r)$ & \\
\\
$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes \\
$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & \\
\\
$v$ & ::= & & values \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values \\
  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple \\
  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address \\
  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring \\
  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & region relationship instance \\
\\
$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
\\
$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
\\
$e$ & ::= & & expressions \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
  &$\mid$& $id$ &  \\
  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
  &$\mid$& $e_1 + e_2$ & integer operations \\
  &$\mid$& $e_1 < e_2$ & comparison operations \\
  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & task calls \\
  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\

@@

\section{Legion Core Language}
\label{sec:legioncore}

\subsection{Language Definition}
\label{subsec:langdef}

@@

\forall i, reduces_{id}(r_i) \in \Phi^*

@@

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \wedge \\
\hspace{1cm} \Gamma(id) = \forall \otok{k}{r}. (\oton{T}), \Phi', Q' \rightarrow T_r \Rightarrow reduces_{id}(r) \in \Phi^*
\end{array}$
}
\caption{Privilege and Constraint Closure}
\label{fig:closure}
\end{figure}

@@

\begin{center}
\begin{tabular}{l@{ }l@{ }l}
$M \vdash E : \Phi$ & $\Leftrightarrow$ & $\forall \epsilon\text { in } E.$ \\
& & $(\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge$ \\
& & $(\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge$ \\
& & $(\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)$
\end{tabular}
\end{center}

@@

\begin{lem}
\label{lemma:phiclosure}
$M \vdash E : \Phi^*$ if and only if $M \vdash E : \Phi$.
\end{lem}

@@

\begin{figure*}
$$
\begin{array}{lll}
valid\_interleave(S, C, E', \oton{E}) & = & any\_interleave(E', \oton{E}) \wedge \\
& & coherent(S, L_{excl}(E', C), L_{excl}(E', C), E')\ \wedge \\
& & seq\_equiv(S, L_{excl}(E', C), L_{excl}(E', C), E', \oton{E})\ \wedge \\
& & coherent(S, L_{atomic}(E', C), \emptyset, E')\ \wedge \\
& & \exists \oton{\pi}, (\oton{\pi}) \text{ is a permutation of } (1,\ldots,n)\ \wedge \\
& & \ \ \ seq\_equiv(S, L_{atomic}(E', C), \emptyset, E', E_{\pi_1}, \ldots, E_{\pi_n}) \\
coherent(S, L_1, L_2, []) & = & \text{true} \\
coherent(S, L_1, L_2, \epsilon \dplus E) & = &\
\begin{cases}
(l \in L_2 \Rightarrow S(l) = v)\ \wedge \\
\ \ \ coherent(S, L_1, L_2, E), & \text{if $\epsilon = read(l, c, v)$} \\
coherent(apply(S, \epsilon), L_1, L_2 \cup \{ l \}, E), & \text{if $\epsilon = write(l, c, v)$ and $l \in L_1$} \\
coherent(apply(S, \epsilon), L_1, L_2, E), & \text{otherwise}
\end{cases} \\
seq\_equiv(S, L_1, L_2, E', \oton{E}) & = & coherent(S, L_1, L_2, E_1 \dplus \ldots \dplus E_2)\ \wedge \\
& & \forall l \in L_1. apply(S, E')(l) = apply(S, E_1 \dplus \ldots \dplus E_n)(l) \\
L_{excl}(E, C) & = & \{ l : op(l, excl, v) \text{ in } E  \} \setminus C \\
L_{atomic}(E, C) & = & \{ l : op(l, atomic, v) \text{ in } E \} \setminus (C \cup L_{excl}(E, C)) \\
\end{array}
$$
\caption{Valid Interleaving Test}
\label{fig:validinterleave}
\end{figure*}

Figure~\ref{fig:validinterleave} gives the precise definition of the {\em valid\_interleave}
predicate, but there are a lot of moving parts, so description in ``plain'' English may help.
Consider a set of constituent traces $\oton{E}$.  Let the {\em exclusive location set} $L_{excl}$ be
the set of all locations that appear in at least one operation with coherence mode $excl$ in at least
one constituent trace, excepting any locations in the clobber set $C$.  Let the {\em atomic location
set} $L_{atomic}$ similarly be the set of locations that appear in at least one operation with 
coherence mode $atomic$, but no operations with coherence mode $excl$, again excepting any location
in the clobber set.  The $coherent(S,L_1,L_2,E)$ test checks that all reads of locations in $L_2$ and any read of a location in $L_1$ that comes after a write to the same location return values
that match what was most recently written into the store.  (The difference betwen $L_1$ and $L_2$
defers the checking of atomic reads that don't have a preceeding write until any other traces that
may be ordered before $E$ are included.)  The $seq\_equiv$ test verifies that every read in $E'$ with a location in $L_1$ would also have be coherent if the operations were performed in the original program order and that the store that results from applying $E'$ matches the store that would
have resulted from applying the constituent traces in program order for all locations in $L_1$.
For the atomic location set, we allow the constituent traces to be permuted, but all locations
must be checked with the same permutation.

@@

To summarize this, we define a {\em noninterference} operator on two effects that have been
annotated with which constituent trace they came from as follows:
$$
\begin{array}{l@{ }l}
op_1^{E_1}(l_1, c_1,& v_1)\ \nonint\ op_2^{E_2}(l_2, c_2, v_2) \Leftrightarrow E_1 \not= E_2\ \wedge \\
& \big( l_1 \not= l_2 \vee (op_1 = read \wedge op_2 = read)\ \vee \\
& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2) \big)
\end{array}
$$

\begin{lem}
\label{lem:nonintswap}
Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \nonint \epsilon_2$.
Then,
$$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$$
if and only if
$$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$$
\end{lem}

\begin{proof}
As before, we consider the three conditions needed to be a valid interleaving.  We describe the
implication in the ``forward'' direction, but the arguments are all symmetric.
\begin{enumerate}
\item If $E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b$ is an interleaving of $\oton{E}$,
the proof must first ``pop off'' all of $E'_a$, leaving $\epsilon_1 \dplus \epsilon_2 \dplus E'_b$ for
the interleaved subtrace.  The next two steps must pop $\epsilon_1$ and $\epsilon_2$, leaving $E'_b$.
They cannot be in the same subtrace because $\epsilon_1 \nonint \epsilon_2$, so we could also pop
$\epsilon_2$ and then $\epsilon_1$, resulting in a proof that $E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b$ is also an interleaving of the constituent traces.
\item For coherence, we must show the value read (if any) by $\epsilon_1$ does not change, the value
read (if any) by $\epsilon_2$ does not change, and that any read in $E'_b$ sees no change.  With the
help of Lemma~\ref{lemma:applychain}, we can eliminate the common $E'_a$ from all sequences by
using $S' = apply(S, E'_a)$.  Similarly to show that $apply(S', \epsilon_1 \dplus \epsilon_2 \dplus E) = apply(S', \epsilon_2 \dplus \epsilon_1 \dplus E)$ for any $E$, we need only to show that it holds
for $E = []$.  The proof is now reduced to showing the following:
\begin{enumerate}
\item $\epsilon_1 = read(l, c, v) \Rightarrow S'(l) = apply(S',\epsilon_2)(l)$ - since $\epsilon_1$
is a read, either $\epsilon_2$ must be a read (in which case $apply(S, \epsilon_2) = S'$) or must be
to a different location, so the result of applying just it to $S'$ cannot change the value in 
location $l$.
\item $\epsilon_2 = read(l, c, v) \Rightarrow apply(S', \epsilon_1)(l) = S'(l)$ - this is the same
as above, with $\epsilon_1$ and $\epsilon_2$ switched.
\item $apply(S', \epsilon_1 \dplus \epsilon_2) = apply(S', \epsilon_2 \dplus \epsilon_1)$ - we must
show that the two store operations $apply(\bullet, \epsilon_1)$ and $apply(\bullet, \epsilon_2)$
commute.  Modifications to two different locations commute, as do two reads (which make no change
to the store).  Finally, two reductions using the same function to the same location also commute.
\end{enumerate}
\item The proof that sequential equivalence is maintained comes down to showing the final store
is identical after the swap of the two operations, and this is merely a special case of the third
piece of the coherence proof.
\end{enumerate}
\end{proof}

We now extend the noninterference operator to entire memory traces, providing a way to say that
no operation in a trace interferes with any operation in other:
$$E_1 \nonint E_2 \Leftrightarrow \displaystyle\bigwedge_{\epsilon_1 \text{ in } E_1, \epsilon_2 \text{ in } E_2} \epsilon_1 \nonint \epsilon_2$$
and observe that this property permits arbitrary ordering (i.e. parallel execution) of the 
corresponding subexpressions.

\begin{lem}
\label{lem:noninteffects}
Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}$ be memory traces such that $E_i \nonint E_j$
for every $1 \leq i < j \leq n$.  Then any trace $E'$ satisfying:
$$any\_interleave(E', \oton{E})$$
is a valid interleaving.
\end{lem}

@@

The swapping algorithm as follows: Associate with each operation in $E'$ the index of that operation
in $E_1 \dplus \ldots \dplus E_n$.  Any consectutive pair of operations $\epsilon_1$ and $\epsilon_2$
for which $index(\epsilon_1) > index(\epsilon_2)$ is a {\em misordered pair}.  For any such pair,
$\epsilon_1$ and $\epsilon_2$ will be from different constituent traces (if they were from the same
trace, their misordering would mean that $E'$ was not an interleaving of $\oton{E}$).  Every pair of
traces is noninterfering, so $\epsilon_1 \nonint \epsilon_2$ and they can be swapped while preserving
validity of the trace.  At most $n^2/2$ swaps are needed to eliminate all misordered pairs.
\end{proof}

Our next step is to extend noninterference to the privileges used in the type system.  We will make
use of the region mapping $M$ to translate the logical regions named in privileges into sets of
locations (i.e. physical regions), and define noninterference between individual privileges as
follows:
$$
\begin{array}{l@{ }l}
priv_1(r_1) \nonint[M] & priv_2(r_2) \Leftrightarrow (M(r_1) \cap M(r_2) = \emptyset)\ \vee \\
& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)
\end{array}
$$
%% \begin{center}
%% \begin{tabular}{llc}
%% \multicolumn{1}{c}{$\phi_1$} & \multicolumn{1}{c}{$\phi_2$} & $\phi_1 \nonint[M] \phi_2$? \\
%% \multirow{3}{*}{$reads(r_1)$} & $reads(r_2)$ & true \\
%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
%% & $reduces_{id_2}(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
%% \cline{1-1}
%% \multirow{3}{*}{$writes(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
%% & $reduces_{id_2}(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
%% \cline{1-1}
%% \multirow{3}{*}{$reduces_{id_1}(r_1)$} & $reads(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
%% & $writes(r_2)$ & $M(r_1) \cap M(r_2) = \emptyset$ \\
%% & $reduces_{id_2}(r_2)$ & $\begin{array}{c}id_1 = id_2 \vee \\ M(r_1) \cap M(r_2) = \emptyset\end{array}$ \\
%% \end{tabular}
%% \end{center}
and noninterference between sets of privileges in the same way as memory traces:
$$\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2$$

@@

\begin{thm}
\label{thm:parallelexec}
Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
Let $M$ be a region mapping, $L$ be local value mapping,
$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
$$
\begin{array}{l}
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$$
If $\Phi_i \nonint[M] \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
$\oton{e}$ will result in a valid interleaving of memory operations.
\end{thm}

@@

\subsection{Relaxed Coherence}
